Isomorphic Interpreters from Logically Reversible Abstract Machines

### 2012-02-24 iso-int/rc-iso-int.tex
sec 2 done

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Simple Bounded Iterator}
We illustrate the main ideas underlying the derivation of isomorphic
interpreters using a simple abstract machine.  Consider the definition
of the following small-step abstract machine:

\begin{multicols}{2}
%subcode{bnf} include main
% Numbers, n, m = 0 | n + 1
% Machine states = <n, n>

@@

The triangle denotes the {{unfold}} isomorphism, which works as
follows: If the number {{n}} is zero, the output is the top branch
which has type {{1}}. If the number was non-zero, the output is on the
bottom branch and has value {{n-1}}.

%% Now numbers can be implemented by the recursive type {{nat = @@x.1+x}}
%% in {{langRevT}}. This gives us the isomorphisms

%% \begin{center}

%% \end{center}

%% Since the abstract machine increments second component, we know that
%% we need a {{fold}} on the second component to accomplish this. So we
%% can fill in the output part of the interpreter dually.

Let us examine all the possible input states, by distributing on the
input. Dually let us do the same with the output, but this time also
be a little bit explicit about the swap operations required so the
right types are in place.

@@

The construction above already gives us some insight into how we can
implement our one reduction step. If we chose {{m = m'-1}} and we
chose {{n-1 =n'}} then we have essentially encoded the required
rewrite. Lets state this in words - the new value of {{n}}, namely
{{n'}}, is equal to {{n-1}}. And further - the new value of {{m}},
namely {{m'}}, is {{m+1}} (because {{m'-1=m}}).

We can thus connect the lower branches with the appropriate
isomorphism to reflect this relationship. In this case the isomorphism
is just a swap of the two wires involved.

@@

We are a few steps away from an interpreter at this point and the
final steps rely on the answers to a few questions.

\begin{enumerate}
\item How do relate the outputs of the combinator to the inputs, such
  that the machine correctly takes multiple steps?

\item

 What do we do about the two inner branches?

The branch labeled {{((), m)}} looks like the stop state of the
machine. The branch label-led {{((), n')}} looks like the start state
of the program. To complete the interpreter, we flip our current
construction inside out.

\begin{center}
  \includegraphics{diagrams/nat-nat6.pdf}
\end{center}

Which gives us

\begin{center}
  \includegraphics{diagrams/nat-nat7.pdf}
\end{center}

This basically completely our construction of the interpreter. We
basically need to use a {{trace}} to feedback related machine states
to related machine states.

\begin{center}
  \includegraphics{diagrams/nat-nat8.pdf}
\end{center}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Simple Bounded Iterator}
We illustrate the main ideas underlying the derivation of isomorphic
interpreters using a simple abstract machine.  Consider the definition
of the following small-step abstract machine:
%\vspace{-30px}
\begin{multicols}{2}
%subcode{bnf} include main
% Numbers, n, m = 0 | n + 1
% Machine states = <n, n>

@@

The triangle on the left denotes the {{unfold}} isomorphism, which
works as follows: If the number {{n}} is zero, the output is the top
branch which has type {{1}}. If the number was non-zero, the output is
on the bottom branch and has value {{n-1}}. The triangle on the right
is the dual {{fold}} isomorphism.  Let us examine all the possible
machine states by introducing distribution operations and any required
swaps.

%% by distributing on the
%% input. Dually let us do the same with the output, but this time also
%% be a little bit explicit about the swap operations required so the
%% right types are in place.

@@

The construction above already gives us some insight into how we can
implement our machine's one reduction step. We want the output {{n'}}
to be {{n+1}} where {{n}} is the first component of the input and
similarly the output {{m'}} to be {{m+1}} where {{m}} is the second
component of the input.  We can express these equalities by connecting
wires -- by connecting the {{n-1}} wire to {{n'}} we express the
equality {{n' =n-1}}. Similarly we connect {{m}} to {{m'-1}} to
express {{m' = m+1}}.

@@

%% We are a few steps away from an interpreter at this point and the
%% final steps rely on the answers to a few questions.

%% \begin{enumerate}
%% \item How do relate the outputs of the combinator to the inputs, such
%%   that the machine correctly takes multiple steps?

%% \item

%%  What do we do about the two inner branches?

%% \end{enumerate}

The branch labeled {{((), m)}} corresponds to the machine state
{{<0,m>}} and hence looks like the stop state of the
machine. Similarly the branch labeled {{((), n')}} corresponds to the
start state of the machine.

We are close to the completion of the interpreter at this point. The
last step in the construction involves introducing a {{trace}]
operation for iteration of reduction steps and exposing the start and
stop states. We do this by `flipping' the construction so far
inside-out.

\vspace{-20px}
\begin{multicols}{2}
\begin{center}
\scalebox{0.5}{
  \includegraphics{diagrams/nat-nat6.pdf}
}
\end{center}
\columnbreak
\begin{center}
\scalebox{0.6}{
  \includegraphics{diagrams/nat-nat7.pdf}
}
\end{center}  
\end{multicols}

The introduced {{trace}} feeds back related machine states. The
exposed start state essentially requires the initial value of {{n}} to
start the machine.

\begin{center}
\scalebox{1.0}{ \includegraphics{diagrams/nat-nat8.pdf} }
\end{center}
